---
title: "Preliminary State Report"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(reshape2)
library(ggplot2)
library(scales)
library(stringr)
knitr::opts_chunk$set(echo = TRUE)
if (!exists("survey")) { 
     survey = read.csv('data.csv')
     state <- 'KY'
     data <- survey[ which(survey$state==state ), ]
}
data$jurisdiction <- sub("^$", "None Specified", data$jurisdiction)
jurisdiction <- group_by(data[ which(data$jurisdiction!=''),], jurisdiction)
jurisdiction <- jurisdiction[ which(jurisdiction$jurisdiction!='None Specified'), ]
```

# State Report for `r state.name[grep(state, state.abb)]`

Of the **`r nrow(data)`** officers in `r state.name[grep(state, state.abb)]` that took this survey, **`r sum(data$num_overdose_calls>0, na.rm=TRUE)`** officers reported having responded to an overdose call within the past six months. This accounted for **`r round(sum(data$num_overdose_calls>0, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers.

The **`r nrow(data)`** officers that responded to this survey reported a total of **`r sum(data$num_overdose_calls, na.rm=TRUE)`** overdose calls within the past six months, with each of the **`r sum(data$num_overdose_calls>0, na.rm=TRUE)`** officers that responded to a call reporting an average of **`r round(sum(data$num_overdose_calls, na.rm=TRUE)/sum(data$num_overdose_calls>0, na.rm=TRUE),2)`** calls per officer.

**`r sum(data$num_naloxone_administer>0, na.rm=TRUE)`** officers in `r state.name[grep(state, state.abb)]` reported having reversed an overdose with naloxone in the past six months, accounting for **`r round(sum(data$num_naloxone_administer>0, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers. There were a total of **`r sum(data$num_naloxone_administer, na.rm=TRUE)`** overdose reversals reported, with each of the **`r sum(data$num_naloxone_administer>0, na.rm=TRUE)`** officers that reported reversing an overdose reporting **`r round(sum(data$num_naloxone_administer, na.rm=TRUE)/sum(data$num_naloxone_administer>0, na.rm=TRUE),2)`** reversals per officer.

```{r echo=FALSE, fig.height=2.5, fig.width=8}
myvars <- c('num_naloxone_administer','num_overdose_calls')
graph <- filter(jurisdiction,n() > 4)
graph <- aggregate(graph[myvars], list(graph$jurisdiction), mean, na.rm = TRUE)
graph <- melt(graph,id.vars="Group.1")
graphlabels <- c("Number of overdoses reversed","Number of overdose calls responded to")

ggplot(graph, aes(variable, value)) +
  geom_bar(aes(fill = Group.1), position = "dodge", stat="identity", color="black") +
  ylab("Average per officer") +
  labs(title = "Overdose Responses Per Officer by Jurisdiction in Last 6 Months", fill = "Jurisdiction", caption = paste("Includes only jurisdictions with five or more officers")) +
  scale_x_discrete(name = "Question", labels = str_wrap(graphlabels, width = 15)) +
  coord_flip() +
  scale_fill_grey()
```

**`r sum(data$narcan_carry>0, na.rm=TRUE)`** officers reported working in departments that equip some or all officers to carry naloxone (this may not mean that every officer is equipped with naloxone). This accounted for **`r round(sum(data$narcan_carry>0, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers in `r state.name[grep(state, state.abb)]`.

```{r echo=FALSE, fig.height=1.5, fig.width=8}
myvars <- c('narcan_carry')
graph <- filter(jurisdiction,n() > 4)
graph <- aggregate(graph[myvars], list(graph$jurisdiction), mean, na.rm = TRUE)
graph <- melt(graph,id.vars="Group.1")
graphlabels <- c("Department equipped to carry narcan")

ggplot(graph, aes(variable, value)) +
  geom_bar(aes(fill = Group.1), position = "dodge", stat="identity", color="black") +
  ylab("Percent of Officers Reporting Departmental Naloxone Use") +
  labs(title = "Naloxone Use by Jurisdiction", fill = "Jurisdiction", caption = paste("Includes only jurisdictions with five or more officers")) +
  scale_y_continuous(labels = percent_format(), limits = c(0,1)) +
  scale_x_discrete(name = "Question", labels = str_wrap(graphlabels, width = 15)) +
  coord_flip() +
  scale_fill_grey()
```

## Officers that responded to overdose calls

The following are results from the **`r sum(data$num_overdose_calls>0, na.rm=TRUE)`** officers in `r state.name[grep(state, state.abb)]` that responded to at least one overdose call in the past 6 months.

Of the **`r sum(data$num_overdose_calls, na.rm=TRUE)`** overdose calls responded to in the past 6 months by **`r sum(data$num_overdose_calls>0, na.rm=TRUE)`** officers in `r state.name[grep(state, state.abb)]`, **`r sum(data$od_arrest, na.rm=TRUE)`** of those calls resulted in the victim of an overdose being arrested. This accounted for **`r round(sum(data$od_arrest, na.rm=TRUE)/sum(data$num_overdose_calls, na.rm=TRUE)*100,2)`%** of all calls. 

Of the **`r sum(data$num_overdose_calls, na.rm=TRUE)`** overdose calls in the past 6 months in `r state.name[grep(state, state.abb)]`, **`r sum(data$overdose_witness_arrest, na.rm=TRUE)`** of those calls resulted in a witness to an overdose being arrested. This accounted for **`r round(sum(data$overdose_witness_arrest, na.rm=TRUE)/sum(data$num_overdose_calls, na.rm=TRUE)*100,2)`%** of all calls. 

\newpage

##Knowledge

Of the **`r nrow(data)`** officers in `r state.name[grep(state, state.abb)]`, **`r sum(data$gsl_na, na.rm=TRUE)`** responded correctly when asked whether `r state.name[grep(state, state.abb)]` has a 911 Good Samaritan Law in effect. This accounted for **`r round(sum(data$gsl_na>0, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers in `r state.name[grep(state, state.abb)]`.

**`r sum(data$gsl_arrest, na.rm=TRUE)`** of the **`r nrow(data)`** officers in `r state.name[grep(state, state.abb)]` responded correctly when asked whether `r state.name[grep(state, state.abb)]` provides immunity from arrest. This accounted for **`r round(sum(data$gsl_arrest>0, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers.

**`r sum(data$gsl_charge, na.rm=TRUE)`** officers responded correctly when asked whether `r state.name[grep(state, state.abb)]` provides immunity from criminal charges. This accounted for **`r round(sum(data$gsl_charge>0, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers in `r state.name[grep(state, state.abb)]`.


```{r echo=FALSE, fig.height=3.5, fig.width=8}
myvars <- c('gsl_charge',	'gsl_arrest',	'gsl_na')
graph <- filter(jurisdiction,n() > 4)
graph <- aggregate(graph[myvars], list(graph$jurisdiction), mean, na.rm = TRUE)
graph <- melt(graph,id.vars="Group.1")
graphlabels <- c("Does your state's GSL protect against charges?","Does your state's GSL protect against arrest?","Does your state have a GSL in effect?")

ggplot(graph, aes(variable, value)) +
  geom_bar(aes(fill = Group.1), position = "dodge", stat="identity", color="black") +
  ylab("Percent of Officers that Answered Correctly") +
  labs(title = "Knowledge of GSL by Jurisdiction", fill = "Jurisdiction", caption = paste("Includes only jurisdictions with five or more officers")) +
  scale_y_continuous(labels = percent_format(), limits = c(0,1)) +
  scale_x_discrete(name = "Question", labels = str_wrap(graphlabels, width = 15)) +
  coord_flip() +
  scale_fill_grey()
```

\newpage

## Training

Of the **`r nrow(data)`** officers in `r state.name[grep(state, state.abb)]`, **`r sum(data$add_training_identify_and_respon>0, na.rm=TRUE)`** stated that they would like additional training in **Identifying and responding to opioid overdoses**. This accounted for **`r round(sum(data$add_training_identify_and_respon>0, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers in `r state.name[grep(state, state.abb)]`.

**`r sum(data$add_training_safe_handling>0, na.rm=TRUE)`** officers (**`r round(sum(data$add_training_safe_handling>0, na.rm=TRUE)/nrow(data)*100,2)`%**) reported they would like additional training in **The safe handling of opioids and other illicit substances in the field**.

**`r sum(data$add_training_science_of_addictio>0, na.rm=TRUE)`** officers (**`r round(sum(data$add_training_science_of_addictio>0, na.rm=TRUE)/nrow(data)*100,2)`%**) reported they would like training in **The clinical science and psychology of opioid use disorder (addiction)**.

**`r sum(data$add_training_procedures_of_gsl>0, na.rm=TRUE)`** officers (**`r round(sum(data$add_training_procedures_of_gsl>0, na.rm=TRUE)/nrow(data)*100,2)`%**) reported they would like training in **Provisions and procedures related to `r state.name[grep(state, state.abb)]` 911 Good Samaritan Law**.

**`r sum(data$add_training_services_available>0, na.rm=TRUE)`** officers (**`r round(sum(data$add_training_services_available>0, na.rm=TRUE)/nrow(data)*100,2)`%**) reported they would like training in **Medical, social, and psychological services available in `r state.name[grep(state, state.abb)]` for people experiencing opioid use disorder (addiction) and their families**.


```{r echo=FALSE, fig.height=6, fig.width=8}
myvars <- c('add_training_services_available', 'add_training_procedures_of_gsl',	'add_training_txt_addiction',	'add_training_science_of_addictio',	'add_training_safe_handling',	'add_training_identify_and_respon')
graph <- filter(jurisdiction,n() > 4)
graph <- aggregate(graph[myvars], list(graph$jurisdiction), mean, na.rm = TRUE)
graph <- melt(graph,id.vars="Group.1")
graphlabels <- c("Treatment Services Available","State GSL Procedures","Treatments for Addiction","Science of Addiction","Safe Handling of Opioids or Substances","Identify and Respond to OD")

ggplot(graph, aes(variable, value)) +
  geom_bar(aes(fill = Group.1), position = "dodge", stat="identity", color="black") +
  ylab("Percent of Officers that Want Training") +
  labs(title = "Training Needs by Jurisdiction", fill = "Jurisdiction", caption = paste("Includes only jurisdictions with five or more officers")) +
  scale_y_continuous(labels = percent_format(), limits = c(0,1)) +
  scale_x_discrete(name = "Type of Training", labels = str_wrap(graphlabels, width = 15)) +
  coord_flip() +
  scale_fill_grey()
```

Of the **`r nrow(data)`** officers in `r state.name[grep(state, state.abb)]`, **`r sum(data$add_training_na>0, na.rm=TRUE)`** reported that the **Training already received is satisfactory**. This accounted for **`r round(sum(data$add_training_na>0, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers in `r state.name[grep(state, state.abb)]`.

### Other training topics suggested

Officers in `r state.name[grep(state, state.abb)]` provided the following qualitative feedback regarding recommended trainings in `r state.name[grep(state, state.abb)]`:

- `r data$add_training_other_specify[data$add_training_other_specify!='']`

\newpage

##Attitudes

The following graph depicts how officers reported the attitude of officers in their department about carrying and administering naloxone:

```{r echo=FALSE, message=FALSE}
library(ggplot2)
graphdata = recode(data$attitude_of_naloxone, 'Overwhelmingly negative'=1,'Generally negative'=2,'Mixed'=3,'Generally positive'=4,'Overwhelmingly positive'=5)
qplot(graphdata) + 
  geom_bar() +
  scale_x_continuous(name = 'Attitude of Department', breaks = 1:5, labels=c('Overwhelmingly negative','Generally negative','Mixed','Generally positive','Overwhelmingly positive')) +
  ylab("Number of Ratings") +
  labs(title = "Departmental Attitude on Naloxone") +
  coord_flip() +
  scale_fill_grey()
```

The following graph depicts how officers responded to the question "How does carrying naloxone affect your ability to protect and serve the public?":

```{r echo=FALSE, message=FALSE}
library(ggplot2)
graphdata = recode(data$naloxone_to_protect, 'Greatly worsens'=1,'Somewhat worsens'=2,'No difference'=3,'Somewhat improves'=4,'Greatly improves'=5)
qplot(graphdata) + 
  geom_bar() +
  scale_x_continuous(name = 'Attitude of Officers', breaks = 1:5, labels=c('Greatly worsens','Somewhat worsens','No difference','Somewhat improves','Greatly improves')) +
  ylab("Number of Ratings") +
  labs(title = "Effect of Naloxone on Ability to Serve") +
  coord_flip() +
  scale_fill_grey()
```

The following reflect how many of the **`r nrow(data)`** officers in `r state.name[grep(state, state.abb)]` AGREED with the following statements:

###911 Good Samaritan Laws

- A 911 Good Samaritan Law increases the number of 911 calls by people who witness an overdose (**`r sum(data$cleaned_gsl_increases_od_calls>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_gsl_increases_od_calls>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)
- 911 Good Samaritan Laws may encourage more opioid use by sending the message that drug use is OK (**`r sum(data$cleaned_gsl_encourage_use>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_gsl_encourage_use>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)
- 911 Good Samaritan Laws save lives (**`r sum(data$cleaned_gsl_saves>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_gsl_saves>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)

###Overdose Response Procedures and Protocols

- It is important for police officers to be present at the scene of an overdose to protect medical personnel (**`r sum(data$cleaned_police_protect_medical>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_police_protect_medical>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)
- It is important for police officers to be present at the scene of an overdose, because they may be the first responder to arrive on the scene with Naloxone and save a life (**`r sum(data$cleaned_police_first_responder>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_police_first_responder>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)
- It is important for police officers to be present at the scene of an overdose to enforce the law (**`r sum(data$cleaned_police_enforce_law>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_police_enforce_law>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)

###Naloxone Training, Carry, and Distribution

- I am worried about the liability of officers who carry and/or administer naloxone (**`r sum(data$cleaned_worried_liability>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_worried_liability>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)
- Programs to train and distribute naloxone to lay community members to treat overdose save lives (**`r sum(data$cleaned_naloxone_training_saves>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_naloxone_training_saves>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)
- Programs to train and distribute naloxone to lay community members may encourage opioid use by sending the message that drug use is OK (**`r sum(data$cleaned_naloxone_encourage_use>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_naloxone_encourage_use>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)
- Training community members how to identify and respond to an overdose may encourage opioid drug use by sending the message that drug use is OK (**`r sum(data$cleaned_training_encourage_use>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_training_encourage_use>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)

###Fentanyl in the Drug Supply

- I am worried about myself and other first responders who are called to assist persons suffering overdose, because there is so much fentanyl in the drug supply (**`r sum(data$cleaned_worried_for_self_fenta>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_worried_for_self_fenta>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)
- I am worried about people who use drugs, because there is so much fentanyl in the drug supply (**`r sum(data$cleaned_worried_for_people_fenta>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_worried_for_people_fenta>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)

###Opioid Misuse and Opioid Use Disorder

- Opioid misuse is a disease that should be dealt with through treatment and support services (**`r sum(data$cleaned_opioid_is_disease>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_opioid_is_disease>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)
- Arrest can play an important role in helping drug users change their behavior (**`r sum(data$cleaned_arrest_changes_behavior>=3, na.rm=TRUE)`**; **`r round(sum(data$cleaned_arrest_changes_behavior>=3, na.rm=TRUE)/nrow(data)*100,2)`%** agree)

\newpage

## Agencies

Officers in `r state.name[grep(state, state.abb)]` provided the following feedback regarding what agencies should be involved in post-overdose outreach efforts to connect the overdose victim with treatment and other services:

- Law Enforcement: **`r sum(data$agencies_law_enforcement)`** out of **`r sum(data$agencies_law_enforcement!='NA')`** officers (**`r round(sum(data$agencies_law_enforcement)/sum(data$agencies_law_enforcement!='NA')*100,2)`%**) reported that **Law Enforcement** should be involved.
- Local Hospitals: **`r sum(data$agencies_local_hospitals)`** out of **`r sum(data$agencies_local_hospitals!='NA')`** officers (**`r round(sum(data$agencies_local_hospitals)/sum(data$agencies_local_hospitals!='NA')*100,2)`%**) believe **Local Hospitals** should be involved.
- Fire Departments: **`r sum(data$agencies_fire_departments)`** out of **`r sum(data$agencies_fire_departments!='NA')`** officers (**`r round(sum(data$agencies_fire_departments)/sum(data$agencies_fire_departments!='NA')*100,2)`%**) believe **Fire Departments** should be involved.
- EMS: **`r sum(data$agencies_ems)`** out of **`r sum(data$agencies_ems!='NA')`** officers (**`r round(sum(data$agencies_ems)/sum(data$agencies_ems!='NA')*100,2)`%**) believe that **EMS** should be involved.
- Public Health Agencies: **`r sum(data$agencies_public_health)`** out of **`r sum(data$agencies_public_health!='NA')`** officers (**`r round(sum(data$agencies_public_health)/sum(data$agencies_public_health!='NA')*100,2)`%**) reported that **Public Health Agencies** should be involved.

\newpage

## Demographics

Of the 2869 patrol officers from 20 states that participated in this survey `r nrow(data)` (`r round((nrow(data)/2869)*100,2)`%) serve in `r state.name[grep(state, state.abb)]`.

- Of the **`r nrow(data)`** officers from `r state.name[grep(state, state.abb)]`, **`r sum(data$female<1, na.rm=TRUE)`** reported their gender as male (**`r round(sum(data$female<1, na.rm=TRUE)/nrow(data)*100,2)`%**) and **`r sum(data$female>0, na.rm=TRUE)`** (**`r round(sum(data$female>0, na.rm=TRUE)/nrow(data)*100,2)`%**) reported their gender as female. The remaining officers did not report their gender.
- The **`r nrow(data)`** officers from `r state.name[grep(state, state.abb)]` reported an average of **`r round(mean(data$years_work, na.rm=TRUE), 1)`** years on the force, with responses ranging from **`r min(data$years_work, na.rm=TRUE)`** to **`r max(data$years_work, na.rm=TRUE)`** years.

\newpage

### Personal Connection to the Opioid Epidemic

**`r sum(data$know_add, na.rm=TRUE)`** out of **`r nrow(data)`** officers in `r state.name[grep(state, state.abb)]` reported that they personally know someone who has lived with a substance use disorder. This constitutes **`r round(sum(data$know_add, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers in `r state.name[grep(state, state.abb)]`.

**`r sum(data$know_od, na.rm=TRUE)`** officers in `r state.name[grep(state, state.abb)]` reported that they know someone who has experienced an overdose. This constitutes **`r round(sum(data$know_od, na.rm=TRUE)/nrow(data)*100,2)`%** of all participating officers in `r state.name[grep(state, state.abb)]`.

```{r echo=FALSE, fig.height=2.5, fig.width=8}
myvars <- c('know_od', 'know_add')
graph <- filter(jurisdiction,n() > 4)
graph <- aggregate(graph[myvars], list(graph$jurisdiction), mean, na.rm = TRUE)
graph <- melt(graph,id.vars="Group.1")
graphlabels <- c("Do you know someone who has experienced an overdose?","Do you know someone who has a substance use disorder?")

ggplot(graph, aes(variable, value)) +
  geom_bar(aes(fill = Group.1), position = "dodge", stat="identity", color="black") +
  ylab("Percent of Officers that Answered Yes") +
  labs(title = "Personal Connection to Epidemic", fill = "Jurisdiction", caption = paste("Includes only jurisdictions with five or more officers")) +
  scale_x_discrete(name = "Question", labels = str_wrap(graphlabels, width = 15)) +
  scale_y_continuous(labels = percent_format(), limits = c(0,1)) +
  coord_flip() +
  scale_fill_grey()
```
